---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# shoredate

<!-- badges: start -->
<!-- badges: end -->

The goal of *shoredate* is to offer methods to shoreline date Stone Age sites located along the Norwegian Skagerrak coast based on their present-day elevation and the trajectory of relative sea-level change.

## Installation

You can install the development version of *shoredate* from [GitHub](https://github.com/isakro/shoredate) with:

```{r setup, warning = FALSE, message = FALSE}
# install.packages("devtools")
devtools::install_github("isakro/shoredate")
```

## Example

This is a basic example outlining how to date a single site by manually specifying the site elevation, and using the default setting for the dating procedure and for plotting the date.

```{r date, warning = FALSE, message = FALSE, fig.width = 3, fig.height = 3}
library(shoredate)

# Create example point using the required coordinate system WGS84 UTM32N (EPSG: 32632).
target_point <- sf::st_sfc(sf::st_point(c(579570, 6582982)), crs = 32632)

target_date <- shoreline_date(site = target_point, elevation = 65)

shoredate_plot(target_date)
```
